### 2024-09-24 outputs/paper/paper.qmd
cleaned folder structure, write final cleaned and combined data set

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Robert Ford
thanks: "Code and data are available at: https://github.com/Ford-Robert/STA304_City-Of-Toronto-Data.git"
date: today
date-format: long??
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

# Data {#sec-data}

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

\newpage

# References

### 2024-09-25 scripts/03-test_data.R
added testing

@@ -1,15 +1,53 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests of Cleaned and Combined TTC data
+# Author: Robert Ford
+# Date: 24 September 2024
+# Contact: robert.ford@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites:
+# Any other information needed?
+# Taken, with few modifications, from Luca Carnegie
+# Check out his Repo here: https://github.com/lcarnegie/TTCTransitAnalysis.git
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+dataset <- read_csv("outputs/data/cleaned_dataset.csv")
 
-#### Test data ####
+# Load necessary libraries
+library(dplyr)
+library(testthat)
+
+# Assume df is your data frame
+df <- dataset
+
+test_that("Data frame has correct columns", {
+  expected_cols <- c("Date", "Day", "vehicle", "Location", "Incident", "Delay")
+  expect_equal(colnames(df), expected_cols)
+})
+
+# Test that there are no missing values
+test_that("No missing values", {
+  expect_equal(sum(is.na(df)), 0)
+})
+
+# Test that the Date column is within the expected range and Delay is non-negative
+test_that("Date and Delay values are within reasonable range", {
+  min_date <- as.Date("2014-01-01")
+  max_date <- as.Date("2024-12-31")
+  
+  # Check that the Date is within the range
+  expect_true(all(df$Date >= min_date & df$Date <= max_date))
+  
+  # Check that Delay is non-negative
+  expect_true(all(df$Delay >= 0))
+})
+
+num_delays_over_1000 <- sum(df$Delay > 1000)
+print(num_delays_over_1000)
+
+# Test that categorical columns have expected levels
+test_that("Categorical columns have expected levels", {
+  expect_true(all(df$Day %in% c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))
+  expect_true(all(df$vehicle %in% c("Subway", "Bus", "Streetcar")))
+})

### 2024-09-25 scripts/02-data_cleaning.R
Fixed cleaning mistake

@@ -18,7 +18,7 @@ raw_data_subway <-read_csv("inputs/data/subway_data.csv")
 #subway_codes <- read_csv("inputs/data/subway-delay-codes.csv")
 
 
-#View(raw_bus_data)
+View(raw_data_bus)
 #View(raw_streetcar_data)
 #View(raw_data_subway)
 #str(raw_bus_data)
@@ -70,34 +70,67 @@ raw_data_bus <- raw_data_bus |> select(Date, Day, vehicle, Location, Incident, D
 #Combine the three datasets 
 combined_data <- bind_rows(raw_data_bus, raw_data_streetcar, raw_data_subway)
 
+
+
+
+unique_incidents <- combined_data %>%
+  group_by(Incident) %>%
+  slice(1) %>%
+  ungroup()
+
+#View(unique_incidents)
+
+
 #Fix some categories
 combined_data <- combined_data |> mutate(
   Incident =
     case_match(
       Incident,
-      "Diversion"  ~ "Diversion", 
-      "Security"  ~ "Security", 
+      "Cleaning" ~ "Cleaning", 
+      "Cleaning - Disinfection" ~ "Cleaning",
       "Cleaning - Unsanitary" ~ "Cleaning",
-      "Emergency Services"  ~ "Emergency Services", 
       "Collision - TTC" ~ "Collision",
-      "Mechanical" ~ "Mechanical",
-      "Operations - Operator" ~ "Operations",
-      "Investigation" ~ "Investigation", 
-      "Utilized Off Route" ~ "Diversion",
-      "General Delay" ~ "General Delay", 
-      "Road Blocked - NON-TTC Collision" ~ "Collision",
-      "Held By" ~ "Held By" , 
-      "Vision" ~ "Vision", 
-      "Operations" ~ "Operations", 
       "Collision - TTC Involved" ~ "Collision",
-      "Late Entering Service" ~ "Late Entering Service", 
-      "Overhead" ~ "Overhead", 
-      "Rail/Switches" ~ "Rail/Switches", 
-      "NA" ~ "N/A"
+      "Diversion" ~ "Diversion",
+      "Emergency Services" ~ "Emergency Services",
+      "General Delay" ~ "General Delay",
+      "Held By" ~ "Held By",
+      "Investigation" ~ "Investigation",
+      "Late" ~ "Late",
+      "Late Entering Service" ~ "Late Entering Service",
+      "Late Entering Service - Mechanical" ~ "Late Entering Service",
[…]

### 2024-09-25 outputs/paper/paper.qmd
added some graphs

@@
 
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+
+library(lubridate)
+library(opendatatoronto)
+library(tidyverse)
+library(tidyr)
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(scales)
+library(gridExtra)
+library(patchwork)
+```
+
 ```{r}
-#| include: false
-#| warning: false
-#| message: false
+data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
 
-library(tidyverse)
-library(palmerpenguins)
 ```
@@
 # Data {#sec-data}
+
+```{r num-incidents-by-mode}
+plot1 <- ggplot(data, aes(x = vehicle)) +
+  geom_bar(fill = "steelblue") +
+  scale_y_continuous(labels = comma) +
+  labs(title = "Number of Incidents by Vehicle Type", x = "Vehicle Type", y = "Number of Incidents") +
+  theme_minimal()
+
+cumulative_delay <- data %>%
+  group_by(vehicle) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE))
+
+plot2 <- ggplot(cumulative_delay, aes(x = vehicle, y = total_delay)) +
+  geom_bar(stat = "identity", fill = "tomato") +
+  scale_y_continuous(labels = comma) +
+  labs(title = "Total Cumulative Delay Time by Vehicle", x = "Vehicle Type", y = "Total Delay (Minutes)") +
+  theme_minimal()
+
+grid.arrange(plot1, plot2, ncol = 2)
+
+```
+
+```{r ave-delay-by-mode}
+
+average_delay <- data %>%
+  group_by(vehicle) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE))
+
+ggplot(average_delay, aes(x = vehicle, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Average Delay Time by Vehicle Type", x = "Vehicle Type", y = "Average Delay (Minutes)") +
+  theme_minimal()
+
+```
+
+```{r ave-delay-by-incident}
+data_incident_avg_delay <- data %>%
+  group_by(Incident) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
+
+# Create the bar chart
+ggplot(data_incident_avg_delay, aes(x = Incident, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
+  labs(title = "Average Delay Time by Incident Type", 
+       x = "Incident Type", y = "Average Delay (Minutes)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+
+data_incident_by_delay <- data %>%
+  group_by(Incident, vehicle) %>%
+  summarise(incident_count = n(), avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  group_by(Incident) %>%
+  mutate(percentage = (incident_count / sum(incident_count)) * 100) %>%
+  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
+
+# Create the stacked bar chart with percentages
+ggplot(data_incident_by_delay, aes(x = Incident, y = avg_delay, fill = vehicle)) +
+  geom_bar(stat = "identity", position = "fill") +  # Stacked bars, normalized to show proportions
+  labs(title = "Average Delay Time by Incident Type and Vehicle (Proportions)",
+       x = "Incident Type", y = "Proportion of Incidents", fill = "Vehicle Type") +
+  scale_y_continuous(labels = scales::percent) +  # Show percentages on y-axis
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+
+```
+```{r}
+data_total_delay <- data %>%
+  group_by(Incident, vehicle) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Step 2: Reorder the Incident types based on total delay for better visualization
+data_total_delay <- data_total_delay %>%
+  group_by(Incident) %>%
+  mutate(Incident = fct_reorder(Incident, total_delay, .desc = TRUE))  # Reorder by total_delay
+
+# Step 3: Create the stacked bar chart
+ggplot(data_total_delay, aes(x = Incident, y = total_delay, fill = vehicle)) +
+  geom_bar(stat = "identity", position = "stack") +  # Stacked bars for total delay time
+  labs(title = "Total Delay Time by Incident Type and Vehicle", 
+       x = "Incident Type", y = "Total Delay Time (Minutes)", fill = "Vehicle Type") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+```
+
+```{r incident-type}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Delay Incidence by Month
+#| fig-align: center
+# Prepare the data: calculate the percentage of incidents by type for each vehicle
+data_incident_percentage <- data %>%
+  group_by(vehicle, Incident) %>%
+  summarise(incident_count = n(), .groups = 'drop') %>%
+  group_by(vehicle) %>%
+  mutate(percentage = (incident_count / sum(incident_count)) * 100)
+
+# Lump incidents making up less than 5% into "Other"
+data_incident_percentage <- data_incident_percentage %>%
+  mutate(Incident = ifelse(percentage < 5, "Other", Incident)) %>%
+  group_by(vehicle, Incident) %>%
+  summarise(incident_count = sum(incident_count), percentage = sum(percentage), .groups = 'drop')
+
+# Function to create doughnut chart for a specific vehicle
+create_doughnut_chart <- function(vehicle_type) {
+  # Filter data for the specific vehicle type
+  vehicle_data <- data_incident_percentage %>%
+    filter(vehicle == vehicle_type)
+
+  # Create the doughnut chart
+  ggplot(vehicle_data, aes(x = 2, y = percentage, fill = Incident)) +
+    geom_bar(stat = "identity", width = 1) +
+    coord_polar(theta = "y") +
+    xlim(0.5, 2.5) +  # Adjust the x-axis limits to create the doughnut shape
+    geom_text(aes(label = paste0(Incident, "\n", round(percentage, 1), "%")), 
+              position = position_stack(vjust = 0.5), size = 3) +
+    labs(title = paste("Incident Distribution for", vehicle_type)) +
+    theme_void() +  # Clean theme for the doughnut chart
+    theme(legend.position = "none")  # Hide the legend for a cleaner look
+}
+
+# Create doughnut charts for each vehicle type
+bus_doughnut <- create_doughnut_chart("Bus")
+streetcar_doughnut <- create_doughnut_chart("Streetcar")
+subway_doughnut <- create_doughnut_chart("Subway")
+
+# Arrange the charts side by side
+bus_doughnut + streetcar_doughnut + subway_doughnut
+
+```
+
+```{r ave-time-delay}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Delay Incidence by Month
+#| fig-align: center
+# Prepare the data: calculate the average delay time for each incident type and vehicle
+data_incident_avg_delay <- data %>%
+  group_by(vehicle, Incident) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Function to create a bar chart for a specific vehicle
+create_bar_chart <- function(vehicle_type) {
+  # Filter data for the specific vehicle type and reorder incidents by average delay
+  vehicle_data <- data_incident_avg_delay %>%
+    filter(vehicle == vehicle_type) %>%
+    mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
+
+  # Create the bar chart with uniform color and spread the x-axis labels
+  ggplot(vehicle_data, aes(x = Incident, y = avg_delay)) +
+    geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
+    labs(title = paste("Average Delay by Incident Type for", vehicle_type),
+         x = "Incident Type", y = "Average Delay (Minutes)") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))  # Rotate labels for readability
+}
+
+# Generate the separate bar charts for each vehicle type
+bus_bar_chart <- create_bar_chart("Bus")
+streetcar_bar_chart <- create_bar_chart("Streetcar")
+subway_bar_chart <- create_bar_chart("Subway")
+
+# Print each graph individually
+print(bus_bar_chart)
+print(streetcar_bar_chart)
+print(subway_bar_chart)
+
+```
+
+```{r incidents-over-time}
+data_inc_num <- data %>%
+  group_by(month = floor_date(Date, "month"), vehicle) %>%
+  summarise(incident_count = n())
+
+# Create the line graph
+ggplot(data_inc_num, aes(x = month, y = incident_count, color = vehicle, group = vehicle)) +
+  geom_line(size = 1) +
+  labs(title = "Number of Incidents Over Time by Vehicle", x = "Time", y = "Number of Incidents", color = "Vehicle Type") +
+  theme_minimal()
+
+data_monthly_delay <- data %>%
+  group_by(month = floor_date(Date, "month"), vehicle) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Create the line graph for monthly total delay times
+ggplot(data_monthly_delay, aes(x = month, y = total_delay, color = vehicle, group = vehicle)) +
+  geom_line(size = 1) +
+  labs(title = "Total Delay Time Over Time by Vehicle (Monthly)", 
+       x = "Time (Month)", y = "Total Delay Time (Minutes)", color = "Vehicle Type") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r}
+data_total_delay <- data %>%
+  group_by(month = floor_date(Date, "month")) %>%  # Group by month
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Create the line chart
+ggplot(data_total_delay, aes(x = month, y = total_delay)) +
+  geom_line(size = 1, color = "steelblue") +  # Single line with color
+  labs(title = "Total Delay Time Over Time", 
+       x = "Time (Month)", y = "Total Delay Time (Minutes)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 

### 2024-09-26 outputs/paper/paper.qmd
added graphs

--- 2024-09-24 outputs/paper/paper.qmd
@@
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 
+library(lubridate)
+library(opendatatoronto)
 library(tidyverse)
-library(palmerpenguins)
+library(tidyr)
+library(dplyr)
+library(knitr)
+library(ggplot2)
+library(scales)
+library(gridExtra)
+library(patchwork)
 ```
 
-# Introduction
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
+
+```
+
+# Introduction{#sec-introduction}
 
 # Data {#sec-data}
+
+```{r num-incidents-by-mode}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: (left) Total Number of Incidents (right) Total Amount of Delay Time
+#| fig-align: center
+
+plot1 <- ggplot(data, aes(x = vehicle)) +
+  geom_bar(fill = "steelblue") +
+  scale_y_continuous(labels = comma) +
+  labs(title = "Number of Incidents by Vehicle Type", x = "Vehicle Type", y = "Number of Incidents") +
+  theme_minimal()
+
+cumulative_delay <- data %>%
+  group_by(vehicle) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE))
+
+plot2 <- ggplot(cumulative_delay, aes(x = vehicle, y = total_delay)) +
+  geom_bar(stat = "identity", fill = "tomato") +
+  scale_y_continuous(labels = comma) +
+  labs(title = "Cumulative Delay Time by Vehicle", x = "Vehicle Type", y = "Total Delay (Minutes)") +
+  theme_minimal()
+
+grid.arrange(plot1, plot2, ncol = 2)
+
+```
+
+```{r ave-delay-by-mode}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time by Vehicle Type
+#| fig-align: center
+
+average_delay <- data %>%
+  group_by(vehicle) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE))
+
+ggplot(average_delay, aes(x = vehicle, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Average Delay Time by Vehicle Type", x = "Vehicle Type", y = "Average Delay (Minutes)") +
+  theme_minimal()
+
+```
+
+```{r ave-delay-by-incident}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time by Incident Type
+#| fig-align: center
+
+data_incident_avg_delay <- data %>%
+  group_by(Incident) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
+
+# Create the bar chart
+ggplot(data_incident_avg_delay, aes(x = Incident, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
+  labs(title = "Average Delay Time by Incident Type", 
+       x = "Incident Type", y = "Average Delay (Minutes)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+
+data_incident_by_delay <- data %>%
+  group_by(Incident, vehicle) %>%
+  summarise(incident_count = n(), avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  group_by(Incident) %>%
+  mutate(percentage = (incident_count / sum(incident_count)) * 100) %>%
+  mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))
+
+```
+
+```{r ave-delay-bus}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time for Buses by, Incident Type
+#| fig-align: center
+
+# Prepare the data: calculate the average delay time for each incident type and vehicle
+data_incident_avg_delay <- data %>%
+  group_by(vehicle, Incident) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Function to create a bar chart for a specific vehicle
+create_bar_chart <- function(vehicle_type) {
+  # Filter data for the specific vehicle type and reorder incidents by average delay
+  vehicle_data <- data_incident_avg_delay %>%
+    filter(vehicle == vehicle_type) %>%
+    mutate(Incident = fct_reorder(Incident, avg_delay, .desc = TRUE))  # Reorder by avg_delay
+
+  # Create the bar chart with uniform color and spread the x-axis labels
+  ggplot(vehicle_data, aes(x = Incident, y = avg_delay)) +
+    geom_bar(stat = "identity", fill = "steelblue") +  # Set a single color for all bars
+    labs(title = paste("Average Delay by Incident Type for", vehicle_type),
+         x = "Incident Type", y = "Average Delay (Minutes)") +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))  # Rotate labels for readability
+}
+
+# Generate the separate bar charts for each vehicle type
+bus_bar_chart <- create_bar_chart("Bus")
+bus_bar_chart
+
+```
+
+```{r ave-delay-streetcar}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time for Streetcars, by Incident Type
+#| fig-align: center
+streetcar_bar_chart <- create_bar_chart("Streetcar")
+streetcar_bar_chart
+```
+
+```{r ave-delay-subway}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time for Subways, by Incident Type
+#| fig-align: center
+
+subway_bar_chart <- create_bar_chart("Subway")
+subway_bar_chart
+
+```
+
+```{r ave-delay-days}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time by Day
+#| fig-align: center
+# Step 1: Extract the day of the week from the Date column and calculate average delay time by day of the week
+data_avg_delay_by_day <- data %>%
+  mutate(day_of_week = wday(Date, label = TRUE, abbr = FALSE)) %>%  # Extract day of the week from Date
+  group_by(day_of_week) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Step 2: Create the bar chart for average delay by day of the week
+ggplot(data_avg_delay_by_day, aes(x = day_of_week, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +  # Vertical bars for average delay
+  labs(title = "Average Delay Time by Day of the Week", 
+       x = "Day of the Week", y = "Average Delay Time (Minutes)") +
+  theme_minimal()
+
+```
+
+```{r ave-delay-months}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time by Month
+#| fig-align: center
+
+# Step 1: Extract the month from the Date column and calculate average delay time by month
+data_avg_delay_by_month <- data %>%
+  mutate(month = month(Date, label = TRUE, abbr = FALSE)) %>%  # Extract month from Date and get full month names
+  group_by(month) %>%
+  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
+
+# Step 2: Reorder the months to display them in chronological order
+data_avg_delay_by_month <- data_avg_delay_by_month %>%
+  mutate(month = factor(month, levels = month.name))  # Ensure months are ordered chronologically
+
+# Step 3: Create the bar chart with switched axes
+ggplot(data_avg_delay_by_month, aes(x = month, y = avg_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +  # Vertical bars with average delay
+  labs(title = "Average Delay Time by Month", 
+       x = "Month", y = "Average Delay Time (Minutes)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))  # Rotate x-axis labels for readability
+```
+
+```{r incidents-over-time}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Number of Incidents Over Time
+#| fig-align: center
+
+data_inc_num <- data %>%
+  group_by(month = floor_date(Date, "month"), vehicle) %>%
+  summarise(incident_count = n())
+
+# Create the line graph
+ggplot(data_inc_num, aes(x = month, y = incident_count, color = vehicle, group = vehicle)) +
+  geom_line(size = 1) +
+  labs(title = "Number of Incidents Over Time by Vehicle", x = "Time", y = "Number of Incidents", color = "Vehicle Type") +
+  theme_minimal()
+
+```
+
+```{r delay-over-time}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Delay Time from 2014 to 2024
+#| fig-align: center
+
+data_monthly_delay <- data %>%
+  group_by(month = floor_date(Date, "month"), vehicle) %>%
+  summarise(average_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
+
+View(data_monthly_delay)
+
+# Create the line graph for monthly average delay times
+ggplot(data_monthly_delay, aes(x = month, y = average_delay, color = vehicle, group = vehicle)) +
+  geom_line(size = 1) +
+  labs(title = "Average Delay Time Over Time by Vehicle (Monthly)", 
+       x = "Time (Month)", y = "Average Delay Time (Minutes)", color = "Vehicle Type") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
 

### 2024-09-26 scripts/02-data_cleaning.R

@@ -72,7 +72,7 @@ combined_data <- bind_rows(raw_data_bus, raw_data_streetcar, raw_data_subway)
 
 
 
-
+# Need to find all of the possible incidents in order to standardize them
 unique_incidents <- combined_data %>%
   group_by(Incident) %>%
   slice(1) %>%
@@ -80,8 +80,7 @@ unique_incidents <- combined_data %>%
 
 #View(unique_incidents)
 
-
-#Fix some categories
+#Fix all incident categories
 combined_data <- combined_data |> mutate(
   Incident =
     case_match(
@@ -116,7 +115,7 @@ combined_data <- combined_data |> mutate(
       "Road Blocked - NON-TTC Collision" ~ "Collision",
       "Roadblock by Collision - Non-TTC" ~ "Collision",
       "Security" ~ "Security",
-      "Securitty" ~ "Security",  # Handling the typo
+      "Securitty" ~ "Security",
       "Utilized Off Route" ~ "Diversion",
       "Utilizing Off Route" ~ "Diversion",
       "Vision" ~ "Vision",
@@ -127,8 +126,8 @@ combined_data <- combined_data |> mutate(
 )
 View(combined_data)
 
-
-#From Test we need to Remove some rows that are corrupted
+# TODO
+# From Test we need to Remove some rows that are corrupted
 # I first remove any rows that contain an NA, though this data may be accurate
 # I want my analysis to focus on the complete record
 

### 2024-09-26 other/unused_graphs.qmd
updated

---
title: "unused cool graphs"
author: "Robert Ford"
format: pdf
editor: visual
---

```{r incident-type}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Delay Incidence by Month
#| fig-align: center
# Prepare the data: calculate the percentage of incidents by type for each vehicle
data_incident_percentage <- data %>%
  group_by(vehicle, Incident) %>%
  summarise(incident_count = n(), .groups = 'drop') %>%
  group_by(vehicle) %>%
  mutate(percentage = (incident_count / sum(incident_count)) * 100)

# Lump incidents making up less than 5% into "Other"
data_incident_percentage <- data_incident_percentage %>%
  mutate(Incident = ifelse(percentage < 5, "Other", Incident)) %>%
  group_by(vehicle, Incident) %>%
  summarise(incident_count = sum(incident_count), percentage = sum(percentage), .groups = 'drop')

# Function to create doughnut chart for a specific vehicle
create_doughnut_chart <- function(vehicle_type) {
  # Filter data for the specific vehicle type
  vehicle_data <- data_incident_percentage %>%
    filter(vehicle == vehicle_type)

  # Create the doughnut chart
  ggplot(vehicle_data, aes(x = 2, y = percentage, fill = Incident)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    xlim(0.5, 2.5) +  # Adjust the x-axis limits to create the doughnut shape
    geom_text(aes(label = paste0(Incident, "\n", round(percentage, 1), "%")), 
              position = position_stack(vjust = 0.5), size = 3) +
    labs(title = paste("Incident Distribution for", vehicle_type)) +
    theme_void() +  # Clean theme for the doughnut chart
    theme(legend.position = "none")  # Hide the legend for a cleaner look
}

# Create doughnut charts for each vehicle type
bus_doughnut <- create_doughnut_chart("Bus")
streetcar_doughnut <- create_doughnut_chart("Streetcar")
subway_doughnut <- create_doughnut_chart("Subway")

# Arrange the charts side by side
bus_doughnut + streetcar_doughnut + subway_doughnut

```

```{r}
# Step 1: Calculate the average delay per incident by time and vehicle type
data_avg_delay_time <- data %>%
  mutate(year_month = as.Date(format(Date, "%Y-%m-01"))) %>%  # Convert date to year-month
  group_by(year_month, vehicle) %>%
  summarise(avg_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')

# Step 2: Create the line graph with 3 lines for each transport mode
ggplot(data_avg_delay_time, aes(x = year_month, y = avg_delay, color = vehicle, group = vehicle)) +
  geom_line(size = 1) +
  labs(title = "Average Delay Time per Incident by Transport Mode (2014 - 2024)", 
       x = "Time (Year-Month)", y = "Average Delay (Minutes)", color = "Transport Mode") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 2024-09-26 outputs/paper/paper.qmd
updated

@@
 
+```{r}
+# Step 1: Calculate total delay time for each incident type
+data_incident_total_delay <- data %>%
+  group_by(Incident) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  mutate(Incident = fct_reorder(Incident, total_delay, .desc = TRUE))  # Reorder by total_delay
+
+# Step 2: Create the bar chart for total delay time by incident type
+ggplot(data_incident_total_delay, aes(x = Incident, y = total_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Total Delay Time by Incident Type", 
+       x = "Incident Type", y = "Total Delay Time (Minutes)") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+
+```
+
 ```{r ave-delay-by-incident}
@@
 streetcar_bar_chart
+```
+
+```{r}
+# Calculate total delay by vehicle and incident
+data_incident_delay <- data %>%
+  group_by(vehicle, Incident) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE)) %>%
+  ungroup()
+
+# Calculate proportions for each incident by vehicle
+data_incident_delay <- data_incident_delay %>%
+  group_by(vehicle) %>%
+  mutate(total_vehicle_delay = sum(total_delay),
+         proportion = total_delay / total_vehicle_delay * 100) %>%
+  ungroup()
+
+# Keep top 5 incidents per vehicle and group the rest as "Other"
+data_top5_incidents <- data_incident_delay %>%
+  group_by(vehicle) %>%
+  arrange(vehicle, desc(proportion)) %>%
+  mutate(rank = row_number()) %>%
+  mutate(Incident = ifelse(rank > 5, "Other", Incident)) %>%
+  group_by(vehicle, Incident) %>%
+  summarise(total_delay = sum(total_delay),
+            proportion = sum(proportion)) %>%
+  ungroup()
+
+incident_colors <- c(
+  "Cleaning" = "#1f77b4",
+  "Collision" = "#ff7f0e",
+  "Diversion" = "#2ca02c",
+  "Emergency Services" = "#d62728",
+  "General Delay" = "#9467bd",
+  "Held By" = "#8c564b",
+  "Investigation" = "#d62728",
+  "Late Entering Service" = "#7f7f7f",
+  "Late Leaving Garage" = "#bcbd22",
+  "Management" = "#17becf",
+  "Mechanical" = "#1f77b4",
+  "Operations" = "#ff7f0e",
+  "Overhead" = "#2ca02c",
+  "Rail/Switches" = "#2ca02c",
+  "Security" = "#d62728",
+  "Vision" = "#8c564b",
+  "Other" = "#7f7f7f",
+  "N/A" = "#7f7f7f"
+)
+
+# Function to create a pie chart for a given vehicle
+create_pie_chart <- function(data, vehicle_type) {
+  data_vehicle <- data %>% filter(vehicle == vehicle_type)
+
+  ggplot(data_vehicle, aes(x = "", y = proportion, fill = Incident)) +
+    geom_bar(width = 1, stat = "identity", color = "white") +
+    coord_polar("y") +
+    labs(title = vehicle_type) +
+    theme_void() +  # Removes background and axis
+    theme(legend.position = "none",  # Removes the legend
+          plot.title = element_text(hjust = 0.5)) + 
+    geom_text(aes(label = paste0(Incident, ": ", round(proportion, 1), "%")), 
+              position = position_stack(vjust = 0.5), size = 3) +
+    scale_fill_manual(values = incident_colors)  # Apply the consistent color palette
+}
+
+# Create pie charts for each vehicle type
+bus_pie <- create_pie_chart(data_top5_incidents, "Bus")
+streetcar_pie <- create_pie_chart(data_top5_incidents, "Streetcar")
+subway_pie <- create_pie_chart(data_top5_incidents, "Subway")
+
+# Arrange the pie charts side by side
+(bus_pie | streetcar_pie | subway_pie) + 
+  plot_annotation(title = "Proportion of Total Delay Time by Incident (Top 5 + Other)")
+
 ```
@@
 
-View(data_monthly_delay)
-
 # Create the line graph for monthly average delay times

### 2024-09-26 scripts/02-data_cleaning.R

@@ -18,7 +18,7 @@ raw_data_subway <-read_csv("inputs/data/subway_data.csv")
 #subway_codes <- read_csv("inputs/data/subway-delay-codes.csv")
 
 
-View(raw_data_bus)
+#View(raw_data_bus)
 #View(raw_streetcar_data)
 #View(raw_data_subway)
 #str(raw_bus_data)
@@ -114,8 +114,8 @@ combined_data <- combined_data |> mutate(
       "Road Block - Non-TTC Collision" ~ "Collision",
       "Road Blocked - NON-TTC Collision" ~ "Collision",
       "Roadblock by Collision - Non-TTC" ~ "Collision",
-      "Security" ~ "Security",
-      "Securitty" ~ "Security",
+      "Security" ~ "Investigation",
+      "Securitty" ~ "Investigation",
       "Utilized Off Route" ~ "Diversion",
       "Utilizing Off Route" ~ "Diversion",
       "Vision" ~ "Vision",
@@ -126,10 +126,16 @@ combined_data <- combined_data |> mutate(
 )
 View(combined_data)
 
+head(combined_data)
 # TODO
 # From Test we need to Remove some rows that are corrupted
-# I first remove any rows that contain an NA, though this data may be accurate
-# I want my analysis to focus on the complete record
+# remove any rows that contain an NA in the Delay column
+# Remove any rows where Delay is negative
+# Remove any rows where Delay Exceeds 1440
+combined_data <- combined_data %>%
+  filter(!is.na(Delay)) %>%
+  filter(Delay >= 0) %>%
+  filter(Delay <= 998)
 
 
 #### Save data ####

### 2024-09-26 scripts/03-test_data.R

@@ -26,9 +26,9 @@ test_that("Data frame has correct columns", {
   expect_equal(colnames(df), expected_cols)
 })
 
-# Test that there are no missing values
-test_that("No missing values", {
-  expect_equal(sum(is.na(df)), 0)
+# Test that there are no missing Delay values
+test_that("No missing values in Delay column", {
+  expect_equal(sum(is.na(df$Delay)), 0)
 })
 
 # Test that the Date column is within the expected range and Delay is non-negative

### 2024-09-26 other/unused_graphs.qmd
cleaned up paper, and removed pie chart

@@
 ```
+
+```{r}
+# Calculate total delay by vehicle and incident
+data_incident_delay <- data %>%
+  group_by(vehicle, Incident) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE)) %>%
+  ungroup()
+
+# Calculate proportions for each incident by vehicle
+data_incident_delay <- data_incident_delay %>%
+  group_by(vehicle) %>%
+  mutate(total_vehicle_delay = sum(total_delay),
+         proportion = total_delay / total_vehicle_delay * 100) %>%
+  ungroup()
+
+# Keep top 5 incidents per vehicle and group the rest as "Other"
+data_top5_incidents <- data_incident_delay %>%
+  group_by(vehicle) %>%
+  arrange(vehicle, desc(proportion)) %>%
+  mutate(rank = row_number()) %>%
+  mutate(Incident = ifelse(rank > 5, "Other", Incident)) %>%
+  group_by(vehicle, Incident) %>%
+  summarise(total_delay = sum(total_delay),
+            proportion = sum(proportion)) %>%
+  ungroup()
+
+incident_colors <- c(
+  "Cleaning" = "#1f77b4",
+  "Collision" = "#ff7f0e",
+  "Diversion" = "#2ca02c",
+  "Emergency Services" = "#d62728",
+  "General Delay" = "#9467bd",
+  "Held By" = "#8c564b",
+  "Investigation" = "#d62728",
+  "Late Entering Service" = "#7f7f7f",
+  "Late Leaving Garage" = "#bcbd22",
+  "Management" = "#17becf",
+  "Mechanical" = "#1f77b4",
+  "Operations" = "#ff7f0e",
+  "Overhead" = "#2ca02c",
+  "Rail/Switches" = "#2ca02c",
+  "Security" = "#d62728",
+  "Vision" = "#8c564b",
+  "Other" = "#7f7f7f",
+  "N/A" = "#7f7f7f"
+)
+
+# Function to create a pie chart for a given vehicle
+create_pie_chart <- function(data, vehicle_type) {
+  data_vehicle <- data %>% filter(vehicle == vehicle_type)
+
+  ggplot(data_vehicle, aes(x = "", y = proportion, fill = Incident)) +
+    geom_bar(width = 1, stat = "identity", color = "white") +
+    coord_polar("y") +
+    labs(title = vehicle_type) +
+    theme_void() +  # Removes background and axis
+    theme(legend.position = "none",  # Removes the legend
+          plot.title = element_text(hjust = 0.5)) + 
+    geom_text(aes(label = paste0(Incident, ": ", round(proportion, 1), "%")), 
+              position = position_stack(vjust = 0.5), size = 3) +
+    scale_fill_manual(values = incident_colors)  # Apply the consistent color palette
+}
+
+# Create pie charts for each vehicle type
+bus_pie <- create_pie_chart(data_top5_incidents, "Bus")
+streetcar_pie <- create_pie_chart(data_top5_incidents, "Streetcar")
+subway_pie <- create_pie_chart(data_top5_incidents, "Subway")
+
+# Arrange the pie charts side by side
+(bus_pie | streetcar_pie | subway_pie) + 
+  plot_annotation(title = "Proportion of Total Delay Time by Incident (Top 5 + Other)")
+
+```

### 2024-09-26 outputs/paper/paper.qmd
cleaned up paper, and removed pie chart

--- 2024-09-26 outputs/paper/paper.qmd
@@
 
+```{r total-delay-incident}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Average Amount of Delay Time for Buses by, Incident Type
+#| fig-align: center
+
+# Step 1: Calculate total delay time for each incident type
+data_incident_total_delay <- data %>%
+  group_by(Incident) %>%
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop') %>%
+  mutate(Incident = fct_reorder(Incident, total_delay, .desc = TRUE))  # Reorder by total_delay
+
+# Step 2: Create the bar chart for total delay time by incident type
+ggplot(data_incident_total_delay, aes(x = Incident, y = total_delay)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Total Delay Time by Incident Type", 
+       x = "Incident Type", y = "Total Delay Time (Minutes)") +
+  scale_y_continuous(labels = comma) +  # Format y-axis labels to be whole numbers with commas
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10))
+
+```
+
 ```{r ave-delay-by-incident}
@@
 ```
-
 ```{r ave-delay-subway}
@@
 ```
+```{r total-delay-time}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: Total Delay Time from 2014 to 2024
+#| fig-align: center
+# Step 1: Convert the Date column to Date format (if it's not already) and group by month and vehicle type
+data_total_delay_time <- data %>%
+  mutate(Date = as.Date(Date)) %>%  # Ensure Date is in Date format
+  group_by(month = floor_date(Date, "month"), vehicle) %>%  # Group by month and vehicle type
+  summarise(total_delay = sum(Delay, na.rm = TRUE), .groups = 'drop')  # Summarize total delay
+
+# Step 2: Create the line chart with 3 lines for each mode of transport
+ggplot(data_total_delay_time, aes(x = month, y = total_delay, color = vehicle, group = vehicle)) +
+  geom_line(size = 1) +  # Create the line chart with separate lines for each vehicle
+  labs(title = "Total Incident Time by Mode of Transport (2014-2024)", 
+       x = "Time", y = "Total Incident Time (Minutes)", color = "Vehicle Type") +
+  theme_minimal() +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Format the x-axis with yearly labels
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
+
+```
 
@@
 #| warning: false
-#| fig-cap: Delay Time from 2014 to 2024
+#| fig-cap: Average Delay Time from 2014 to 2024
 #| fig-align: center
@@
   summarise(average_delay = mean(Delay, na.rm = TRUE), .groups = 'drop')
-
-View(data_monthly_delay)
 

### 2024-09-27 outputs/paper/paper.qmd
added abstract

@@
 thanks: "Code and data are available at: https://github.com/Ford-Robert/STA304_City-Of-Toronto-Data.git"
-date: today
-date-format: long??
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+date: 27/09/2024
+date-format: 27 September 2024
+abstract: "This study analyzes ten years of delay data from the Toronto Transit Commission (TTC), focusing on subways, streetcars, and buses to assess changes in delay times of these services.
+The analysis found that buses experience the highest number of delays and the longest total and average delay times, while subway delays are resolved much faster on average.
+The analysis also delves into the specific challenges each service faces, analyzing which incidents they are most likely to experience and how long those incidents take to resolve.
+These findings provide data-driven insights for improving TTC operations and uncovering the specific challenges faced by each transportation mode."
 format: pdf

### 2024-09-27 outputs/paper/paper.qmd
added intro

@@
 # Introduction{#sec-introduction}
+Effective public transportation is vital to any large dense urban area.
+It has the capacity to significantly reduce traffic congestion [REF] and improve transportation times for all residents of a city.
+Public transportation also provides affordable transportation options allowing for a more equitable society, in which everyone has access to opportunities across a city [REF].
+Expanding TTC services, and ensuring reliability, is the best option for taking cars off the road which would directly reduce both emissions [REF] and car related deaths within the city [REF].
+
+Despite the importance of the TTC, public opinion in Toronto suggests that the TTC is failing to meet the needs of its riders.
+Complaints about aging infrastructure, frequent delays, and most glaringly safety concerns are widely shared among Torontonians [REF].
+Another point of criticism is that the TTC, especially the subway, has failed to expand much in recent years, despite Toronto growing significantly in population.
+
+Considering these concerns this paper analyzes TTC delay data from 2014 to 2024, to assess whether its service have truly deteriorated over the past decade.
+By examining delays across subways, streetcars, and buses, this study will first compare these services with each other to explore which suffers the most from high delay times and large numbers of incidents.
+Then this study aims to identify the specific challenges each service faces.
+Finally, this paper will explore how delay times and incident rates are distributed over days of the week, and how these metrics have trended over the last ten years.
+
+These findings offer insights for policymakers, transit authorities, and the public.
+By taking a deep dive into the publicly available data, this analysis provides a foundation for understanding TTC delays across buses, streetcars, and subways.
+The paper includes an exploration of the data by constructing easy-to-read graphs, designed to highlight the areas of greatest concern within the TTC.
 

### 2024-09-27 outputs/paper/paper.qmd
added bulk of text

@@
 # Data {#sec-data}
+Delay Data detailing the operations of the TTC was collected and cleaned from the Open Data Toronto [REF] website.
+This data spans about 10 years, from 2014 to 2024.
+Three separate datasets were used, Delay data related to the Subway[REF], Streetcar[REF] and Bus[REF].
+The original datasets are separated into a variety of formats depending on which year you are examining.
+The way in which variable names and the data stored changes over time.
+To further complicate the collection process, the information found in each dataset was different, though it mostly represented the same concepts.
+Data that was too difficult to integrate into one combined dataset that incorporated all ten years, and all three modes of transport was excluded.
+The final dataset contains 1,042,564 observations of 6 variables;
+
+  -Date: The Year, Month and Day that a delay occurred
+
+  -Day: What day of the week that a delay occurred
+
+  -Vehicle: What mode of transport did the delay occur on
+
+  -Location: Where did the delay occur. For Subways Location is the nearest station. For buses and                 streetcars Location is the nearest major intersection or passenger stop.
+
+  -Incident: A description of what caused the delay
+
+  -Delay: How long in minutes the delay lasted
+
+```{r data-table}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-cap: (left) Total Number of Incidents (right) Total Amount of Delay Time
+#| fig-align: center
+
+```
+
+In the cleaning phase, all data entries that had an incomplete (NA) value for the Delay variable were removed.
+Because if a delay occurred but we do not how impactful that delay was, then it does not contribute to the analysis.
+Data that was clearly erroneous was removed, like negative delay times and extremely high delay times.
+Furthermore, all observations whose delay time exceeded 998 minutes (999 minutes was the code for a cancellation between 2015 and 2019) were removed.
 
@@
 ```
+First this paper will assess which Mode of Transport suffers the most from delays.
+From [FIG], we observe that buses suffer the highest number of delays and are delayed for the longest period.
+Interestingly, though the subways suffer from more Incidents than the streetcar, its total time lost to delays is significantly less.
+This would indicate that individual Subway incidents are resolved faster than Streetcar incidents on average.
 
@@
 ```
+
+[FIG2] would support this claim, as on average Streetcar delays take 13 minutes to resolve whereas Subway delays only take about 2.30 minutes to resolve.
+Unfortunately, Buses not only have the most delays, but its delays on average take about 20 minutes to resolve.
+
+NEXT SUBSECT
+To discover why Bus delays are so much worse than Subway and Streetcar delays, we need to observer the differences in how long every type of incident takes to resolve, as well as what type of incident each mode of transport suffers from the most.
+This will also provide crucial insight into which Incidents are the most problematic, then the TTC may focus their efforts on the most disruptive incidents.
 
@@
 ```
+From [FIG3] we can see that Diversions make up the bulk of the total time lost to delays over the last 10 years.
+This could be due to the sheer number of Diversions that happen, or each Diversion causes a large delay.
+By investigating [FIG4] we observe that Diversions on average take a long time to resolve.
 
@@
 ```
+Next, we will break down average delay time by each mode of transport, as this will give us a better understanding of incidents are the most disruptive for each mode of transport.
+In [Fig5] we observe that buses are most impacted by Held by, Overhead, and Diversions.
+To most improve Bus services the TTC should focus on reducing the average delay times of these Incidents.
+[Fig6] displays that the Streetcar is plagued by the same high-cost average delay time incidents.
+However, the spread for Streetcar is broader so targeting these three incidents would not be as impactful as it is for buses.
+However, [Fig7] shows part of why the subway suffers so much less from delays.
+The incidents the subway suffers from take only 4 minutes on average to resolve, in the worst case.
 
@@
 
+NEXT SUBSECT -Delays Over Time-
+
+To further investigate the data, we will investigate when incidents are likely to occur.
+From Fig[] we observe that the average delay time of incidents is about 2 minutes longer on the weekends.
+This could be because there are less TTC staff on the weekends, so incidents take longer to be delt with.
+Across months, FIG, there is little difference in the average delay time.
+This is surprising, as in the winter delays could be worse as weather conditions worsen.
+This may be a testament to the TTC’s ability to deal with Toronto’s harsh winters.
 ```{r ave-delay-days}
@@
 
+Finally, to get a wider view of how incidents have changed over the past 10 years we look to FIG8 and FIG9.
+In Fig8 we see that the number of bus incidents decreased dramatically between 2014 and 2021.
+Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in FIG9.
+This may be due to the TTC changing how they classify incidents, so the data is entered differently but the amount of delay time remains stable.
 ```{r incidents-over-time}
@@
 ```
+
 ```{r total-delay-time}
@@
 ```
+Fig9 suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop.
+This drop could be related to the Covid-19 pandemic for several reasons.
+Firstly, ridership dropped dramatically during the pandemic therefore fewer buses were scheduled and thus fewer Diversions, Overhead and Collisions.
+There is a modest drop in Streetcar Delay times as well which would support this narrative.
+Secondly, the city made public transportation free during the pandemic, so delays due to Investigations would fall.
+In Fig10 observe that average delay time did not noticeably drop during the pandemic.
+This lends further support to these claims, as average Incident delay time remained the same, however far fewer incidents were occurring as the system was constricted.
+Furthermore, Fig10 indicates that over the last 10 years the average delay time per incident has been increasing slightly.
+This could be an indication that the aging fleet of TTC vehicles requires more time to resolve certain incidents.
 
@@
 
-## First discussion point {#sec-first-point}
+The results show that buses experience the highest number of delays and the longest total and average delay times.
+Incidents that have long average delay times such as diversions, mechanical issues, and “held by” incidents contribute most to the total delays buses suffer from.
+In contrast, subway delays are resolved more quickly on average, which may be attributed to less exposure to external factors like traffic, and a more controlled environment allowing for faster repair or replacement.
 
-## Second discussion point
+These findings have important implications for the TTC and its efforts to improve service reliability.
+For example, reducing the frequency and duration of diversions and mechanical issues could have a disproportionate effect on improving bus services.
+Expanding the Subway would also be a great strategy as it seems subway services are inherently more reliable, so maximizing ridership there would be really beneficial.
 
-## Third discussion point
-
-## Weaknesses and next steps
+To build on these findings, further exploration focusing on location data could provide deeper insights into delay patterns across the city.
+Creating a heat map of Toronto that visualizes delay times would identify geographic hotspots where delays are most frequent or severe.
+This spatial analysis could help TTC officials focus their efforts on certain neighborhoods, intersections, or corridors that are more prone to delays.
 

### 2024-09-27 outputs/paper/paper.qmd
cleaned doc, added figure hyperlinks, added bib, added data subsections

@@
 # Data {#sec-data}
+
 Delay Data detailing the operations of the TTC was collected and cleaned from the Open Data Toronto [REF] website.
@@
 #| warning: false
-#| fig-cap: (left) Total Number of Incidents (right) Total Amount of Delay Time
+#| fig-cap: First five rows of the Final Dataset
 #| fig-align: center
 
+kable(head(data, 5), caption = "Sample of the Data")
 ```
@@
 
-```{r num-incidents-by-mode}
+## Delay of Vehicles {#subsec1-data}
+
+First this paper will assess which Mode of Transport suffers the most from delays.
+From @fig-incidents-mode, we observe that buses suffer the highest number of delays and are delayed for the longest period.
+Interestingly, though the subways suffer from more Incidents than the streetcar, its total time lost to delays is significantly less.
+This would indicate that individual Subway incidents are resolved faster than Streetcar incidents on average.
+```{r fig-incidents-mode}
 #| message: false
@@
 ```
-First this paper will assess which Mode of Transport suffers the most from delays.
-From [FIG], we observe that buses suffer the highest number of delays and are delayed for the longest period.
-Interestingly, though the subways suffer from more Incidents than the streetcar, its total time lost to delays is significantly less.
-This would indicate that individual Subway incidents are resolved faster than Streetcar incidents on average.
 
-```{r ave-delay-by-mode}
+```{r fig-ave-delay}
 #| message: false
@@
 
-[FIG2] would support this claim, as on average Streetcar delays take 13 minutes to resolve whereas Subway delays only take about 2.30 minutes to resolve.
+@fig-ave-delay would support this claim, as on average Streetcar delays take 13 minutes to resolve whereas Subway delays only take about 2.30 minutes to resolve.
 Unfortunately, Buses not only have the most delays, but its delays on average take about 20 minutes to resolve.
 
-NEXT SUBSECT
+## Causes of Delays {#subsec2-data}
 To discover why Bus delays are so much worse than Subway and Streetcar delays, we need to observer the differences in how long every type of incident takes to resolve, as well as what type of incident each mode of transport suffers from the most.
@@
 
-```{r total-delay-incident}
+```{r fig-total-delay}
 #| message: false
@@
 ```
-From [FIG3] we can see that Diversions make up the bulk of the total time lost to delays over the last 10 years.
+From @fig-total-delay we can see that Diversions make up the bulk of the total time lost to delays over the last 10 years.
 This could be due to the sheer number of Diversions that happen, or each Diversion causes a large delay.
-By investigating [FIG4] we observe that Diversions on average take a long time to resolve.
+By investigating @fig-delay-incident we observe that Diversions on average take a long time to resolve.
 
-```{r ave-delay-by-incident}
+```{r fig-delay-incident}
 #| message: false
@@
 Next, we will break down average delay time by each mode of transport, as this will give us a better understanding of incidents are the most disruptive for each mode of transport.
-In [Fig5] we observe that buses are most impacted by Held by, Overhead, and Diversions.
+In @fig-delay-bus we observe that buses are most impacted by Held by, Overhead, and Diversions.
 To most improve Bus services the TTC should focus on reducing the average delay times of these Incidents.
-[Fig6] displays that the Streetcar is plagued by the same high-cost average delay time incidents.
+@fig-delay-streetcar displays that the Streetcar is plagued by the same high-cost average delay time incidents.
 However, the spread for Streetcar is broader so targeting these three incidents would not be as impactful as it is for buses.
-However, [Fig7] shows part of why the subway suffers so much less from delays.
+However, @fig-delay-subway shows part of why the subway suffers so much less from delays.
 The incidents the subway suffers from take only 4 minutes on average to resolve, in the worst case.
 
-```{r ave-delay-bus}
+```{r fig-delay-bus}
 #| message: false
@@
 ```
-
-```{r ave-delay-streetcar}
+```{r fig-delay-streetcar}
 #| message: false
@@
 ```
-```{r ave-delay-subway}
+```{r fig-delay-subway}
 #| message: false
@@
 
-NEXT SUBSECT -Delays Over Time-
+## Delays Over Time {#subsec3-data}
 
 To further investigate the data, we will investigate when incidents are likely to occur.
-From Fig[] we observe that the average delay time of incidents is about 2 minutes longer on the weekends.
+From @fig-delay-days we observe that the average delay time of incidents is about 2 minutes longer on the weekends.
 This could be because there are less TTC staff on the weekends, so incidents take longer to be delt with.
-Across months, FIG, there is little difference in the average delay time.
+Across months, @fig-delay-months, there is little difference in the average delay time.
 This is surprising, as in the winter delays could be worse as weather conditions worsen.
 This may be a testament to the TTC’s ability to deal with Toronto’s harsh winters.
-```{r ave-delay-days}
+```{r fig-delay-days}
 #| message: false
@@
 
-```{r ave-delay-months}
+```{r fig-delay-months}
 #| message: false
@@
 
-Finally, to get a wider view of how incidents have changed over the past 10 years we look to FIG8 and FIG9.
-In Fig8 we see that the number of bus incidents decreased dramatically between 2014 and 2021.
-Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in FIG9.
+Finally, to get a wider view of how incidents have changed over the past 10 years we look to @fig-over-time and @fig-total-delay In @fig-over-time we see that the number of bus incidents decreased dramatically between 2014 and 2021.
+Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in @fig-total-delay.
 This may be due to the TTC changing how they classify incidents, so the data is entered differently but the amount of delay time remains stable.
-```{r incidents-over-time}
+```{r fig-over-time}
 #| message: false
@@
 
-```{r total-delay-time}
+```{r fig-total-delay}
 #| message: false
@@
 ```
-Fig9 suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop.
+@fig-total-delay suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop.
 This drop could be related to the Covid-19 pandemic for several reasons.
@@
 Secondly, the city made public transportation free during the pandemic, so delays due to Investigations would fall.
-In Fig10 observe that average delay time did not noticeably drop during the pandemic.
+In @fig-ave-delaytime observe that average delay time did not noticeably drop during the pandemic.
 This lends further support to these claims, as average Incident delay time remained the same, however far fewer incidents were occurring as the system was constricted.
@@
 
-```{r delay-over-time}
+```{r fig-ave-delaytime}
 #| message: false
@@
 For example, reducing the frequency and duration of diversions and mechanical issues could have a disproportionate effect on improving bus services.
-Expanding the Subway would also be a great strategy as it seems subway services are inherently more reliable, so maximizing ridership there would be really beneficial.
+Expanding the Subway would be a great strategy as subway services are inherently more reliable, so maximizing ridership there would be beneficial.
 
@@
 # References
+
+::: {#refs}
+:::

### 2024-09-27 outputs/paper/paper.qmd
Added references. Added in-text citations. Fixed bib. cleaned up file

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
-author: 
-  - Robert Ford
+title: "Ten Years of TTC Delays, an Analysis"
+author: Robert Ford
 thanks: "Code and data are available at: https://github.com/Ford-Robert/STA304_City-Of-Toronto-Data.git"
@@
 Effective public transportation is vital to any large dense urban area.
-It has the capacity to significantly reduce traffic congestion [REF] and improve transportation times for all residents of a city.
-Public transportation also provides affordable transportation options allowing for a more equitable society, in which everyone has access to opportunities across a city [REF].
-Expanding TTC services, and ensuring reliability, is the best option for taking cars off the road which would directly reduce both emissions [REF] and car related deaths within the city [REF].
+It has the capacity to significantly reduce traffic congestion [@anderson] and improve transportation times for all residents of a city.
+Public transportation also provides affordable transportation options allowing for a more equitable society, in which everyone has access to opportunities across a city [@wbpublic].
+Expanding TTC services, and ensuring reliability, is the best option for taking cars off the road which would directly reduce both emissions [@welle] and car related deaths within the city [@buch].
 
 Despite the importance of the TTC, public opinion in Toronto suggests that the TTC is failing to meet the needs of its riders.
-Complaints about aging infrastructure, frequent delays, and most glaringly safety concerns are widely shared among Torontonians [REF].
+Complaints about aging infrastructure, frequent delays, and most glaringly safety concerns are widely shared among Torontonians [@blogto].
 Another point of criticism is that the TTC, especially the subway, has failed to expand much in recent years, despite Toronto growing significantly in population.
@@
 
-Delay Data detailing the operations of the TTC was collected and cleaned from the Open Data Toronto [REF] website.
+Delay Data detailing the operations of the TTC was collected and cleaned from the Open Data Toronto [@citedatatoronto] website.
 This data spans about 10 years, from 2014 to 2024.
-Three separate datasets were used, Delay data related to the Subway[REF], Streetcar[REF] and Bus[REF].
+Three separate datasets were used, Delay data related to the Subway [@subwaydata], Streetcar [@streetcardata] and Bus[@busdata].
 The original datasets are separated into a variety of formats depending on which year you are examining.
@@
 
-Finally, to get a wider view of how incidents have changed over the past 10 years we look to @fig-over-time and @fig-total-delay In @fig-over-time we see that the number of bus incidents decreased dramatically between 2014 and 2021.
-Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in @fig-total-delay.
+Finally, to get a wider view of how incidents have changed over the past 10 years we look to @fig-over-time and @fig-totaltime-delay In @fig-over-time we see that the number of bus incidents decreased dramatically between 2014 and 2021.
+Interestingly this trend of lower number of incidents does not correlate with total incident time, as we observe in @fig-totaltime-delay.
 This may be due to the TTC changing how they classify incidents, so the data is entered differently but the amount of delay time remains stable.
@@
 
-```{r fig-total-delay}
+```{r fig-totaltime-delay}
 #| message: false
@@
 ```
-@fig-total-delay suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop.
+@fig-totaltime-delay suggests that total incident time remains stable for buses, except in 2020 and 2021 where there is a noticeable drop.
 This drop could be related to the Covid-19 pandemic for several reasons.
@@
 # References
-
-::: {#refs}
-:::

### 2024-09-27 scripts/02-data_cleaning.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans the raw bus, streetcar, and subway data
 # Author: Robert Ford
-# Date: 21 September
+# Date: 21 September 2024
 # Contact: robert.ford@mail.utoronto.ca
 # License: MIT
 # Pre-requisites:

### 2024-09-27 scripts/00-simulate_data.R
updated simulation script

@@ -9,11 +9,24 @@
 
 
 #### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
+library(dplyr)
+library(lubridate)
 
 #### Simulate data ####
 set.seed(304)
+
+incident_types <- c('Mechanical', 'General Delay', 'Emergency Services', 'Investigation', 'Signal Issue', 'Medical Emergency')
+vehicles <- c('Bus', 'Streetcar', 'Subway')
+
+# Simulate data
+data_sim <- tibble(
+  Incident = sample(incident_types, 50, replace = TRUE),
+  Delay = sample(5:60, 50, replace = TRUE),
+  Vehicle = sample(vehicles, 50, replace = TRUE),         
+  Date = sample(seq(as.Date('2014-01-01'), as.Date('2024-12-31'), by="day"), 50, replace = TRUE)  # Random dates between 2014 and 2024
+)
+
+print(head(data_sim, 5))
 
 
 

### 2024-09-27 test.R
cleaned up files

@@ -1,47 +1,0 @@
-```{r incident-type}
-#| message: false
-#| echo: false
-#| warning: false
-#| fig-cap: Delay Incidence by Month
-#| fig-align: center
-# Prepare the data: calculate the percentage of incidents by type for each vehicle
-data_incident_percentage <- data %>%
-  group_by(vehicle, Incident) %>%
-  summarise(incident_count = n(), .groups = 'drop') %>%
-  group_by(vehicle) %>%
-  mutate(percentage = (incident_count / sum(incident_count)) * 100)
-
-# Lump incidents making up less than 5% into "Other"
-data_incident_percentage <- data_incident_percentage %>%
-  mutate(Incident = ifelse(percentage < 5, "Other", Incident)) %>%
-  group_by(vehicle, Incident) %>%
-  summarise(incident_count = sum(incident_count), percentage = sum(percentage), .groups = 'drop')
-
-# Function to create doughnut chart for a specific vehicle
-create_doughnut_chart <- function(vehicle_type) {
-  # Filter data for the specific vehicle type
-  vehicle_data <- data_incident_percentage %>%
-    filter(vehicle == vehicle_type)
-  
-  # Create the doughnut chart
-  ggplot(vehicle_data, aes(x = 2, y = percentage, fill = Incident)) +
-    geom_bar(stat = "identity", width = 1) +
-    coord_polar(theta = "y") +
-    xlim(0.5, 2.5) +  # Adjust the x-axis limits to create the doughnut shape
-    geom_text(aes(label = paste0(Incident, "\n", round(percentage, 1), "%")), 
-              position = position_stack(vjust = 0.5), size = 3) +
-    labs(title = paste("Incident Distribution for", vehicle_type)) +
-    theme_void() +  # Clean theme for the doughnut chart
-    theme(legend.position = "none")  # Hide the legend for a cleaner look
-}
-
-# Create doughnut charts for each vehicle type
-bus_doughnut <- create_doughnut_chart("Bus")
-streetcar_doughnut <- create_doughnut_chart("Streetcar")
-subway_doughnut <- create_doughnut_chart("Subway")
-
-# Arrange the charts side by side
-bus_doughnut + streetcar_doughnut + subway_doughnut
-
-
-```

### 2024-09-27 outputs/paper/paper.qmd
fixed stuff

@@
 ---
-title: "Ten Years of TTC Delays, an Analysis"
+title: "Ten Years of TTC Delays an Analysis"
 author: Robert Ford
@@
 The incidents the subway suffers from take only 4 minutes on average to resolve, in the worst case.
-
 ```{r fig-delay-bus}

### 2024-09-27 outputs/paper/paper.qmd
fix

@@
 ```
-```{r fig-delay-streetcar}
+```{r fig-delay-streetcar, fig.pos='H'}
 #| message: false
@@
 ```
-```{r fig-delay-subway}
+```{r fig-delay-subway, fig.pos='H'}
 #| message: false
@@
 
+\newpage
+
 # Discussion {#sec-discussion}

### 2024-09-27 outputs/paper/paper.qmd
fix refs

@@
 date: 27/09/2024
-date-format: 27 September 2024
+date-format: 27 Sept 2024
 abstract: "This study analyzes ten years of delay data from the Toronto Transit Commission (TTC), focusing on subways, streetcars, and buses to assess changes in delay times of these services.
@@
 
+\newpage
+
 # Introduction{#sec-introduction}
@@
 
+\newpage
+
 # Data {#sec-data}
+
+All data was collected and cleaned using R (@citeR), and its tidyverse (@citetidyverse), lubridate (@lubridate), and readxl (@citereadxl) packages.
 
